---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
```{r echo=TRUE}
data <- read.csv("activity.csv")
head(data);
dim <- dim(data)
summary(data)
```
First few rows of data and descriptive summary of its variables shown above. There are `r dim[1]` records in the dataset.

## What is mean total number of steps taken per day?
```{r echo=TRUE}
total_steps_day <- with(data, aggregate(steps ~ date,FUN=sum))
mean_total_steps <- round(mean(total_steps_day$steps))
median_total_steps <- round(median(total_steps_day$steps))

hist(total_steps_day$steps,labels=TRUE,ylim=c(0,30),main="Total number of daily steps",xlab="Number of steps per day")
abline(v=mean_total_steps,col='red',lty=5,lwd=2)
abline(v=median_total_steps,col='blue',lty=3,lwd=4)
```

Mean (red dashed line) of the total number of daily steps was `r format(mean_total_steps,scientific=F)`.
Median (blue dotted line) of the total number of daily steps was `r format(median_total_steps,scientific=F)`.

## What is the average daily activity pattern?
```{r echo=TRUE}
steps_per_timepoint <- with(data, aggregate(steps ~ interval,FUN=mean))

plot(x=steps_per_timepoint$interval,y=steps_per_timepoint$steps,type='l',ylab="Mean",
     main="Average number of steps taken across all days",xlab="Interval")

# Interval with highest average of steps
max_steps <- max(steps_per_timepoint$steps)
interval_max <- steps_per_timepoint$interval[which(steps_per_timepoint$steps==max_steps)]
interval_max

axislabel <- paste0("Max at \n",interval_max)
max_avg_steps <- paste0("Max is ~ ",round(max_steps))
axis(side=1,at=c(interval_max),labels=axislabel,tick=FALSE)
text(x=interval_max,y=max_steps,col="purple",labels=max_avg_steps,pos=4)
```

The 5-minute interval at `r interval_max` contains the maximum number of `r round(max_steps,2)` steps taken, averaged across all days.

## Imputing missing values
```{r echo=TRUE}
miss_vec <- is.na(data$steps)
missings <- sum(miss_vec)
```
There are in total `r missings` missing records in the dataset.

Strategy to replace NAs: sample randomly from all non-NA 5min-interval averages 
```{r echo =TRUE}
library(dplyr)
set.seed(19)
samplev <- sample(x=steps_per_timepoint$steps[!is.na(steps_per_timepoint$steps)],size=dim(data)[1],replace=TRUE)

data_sample <- cbind(data,samplev)
data_imputed <- data_sample %>% mutate(steps = ifelse(is.na(steps),samplev,steps))
```
Reproduce the histogram and averages:
```{r echo=TRUE}
total_steps_day <- with(data_imputed, aggregate(steps ~ date,FUN=sum))
mean_total_steps <- round(mean(total_steps_day$steps))
median_total_steps <- round(median(total_steps_day$steps))

hist(total_steps_day$steps,labels=TRUE,ylim=c(0,40),main="Total number of daily steps (NAs imputed)",
     xlab="Number of steps per day")
abline(v=mean_total_steps,col='red',lty=5,lwd=2)
abline(v=median_total_steps,col='blue',lty=3,lwd=4)
```

Mean (red dashed line) of the total number of daily steps was `r format(mean_total_steps,scientific=F)`.
Median (blue dotted line) of the total number of daily steps was `r format(median_total_steps,scientific=F)`.

## Are there differences in activity patterns between weekdays and weekends?
